Texture mapping 3D planar models of indoor environments with noisy camera poses

### 2013-01-19 spie/paper.tex
everything except geom stuff draft

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\begin{abstract}
  Automated 3D modeling of building interiors is used in applications
  such as virtual reality and environment mapping. Texturing these
  models allows for photorealistic visualizations of the data
  collected by such modeling systems. Camera poses obtained by these
  systems and used for texturing often suffer from inaccuracies
  however, resulting in visible discontinuities when successive images
  are projected onto a surface for texturing. Existing methods to
  stitch images together are often computationally expensive and work
  independently of pose estimates and geometry information. We propose
  an efficient method to refine camera poses in 2D using both existing
  estimates and geometry information, followed by two different
  methods to composite images together, based on the geometry of
  surfaces being textured. The effectiveness of our methods is
  demonstrated on a number of different indoor environments.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:introduction} % \label{} allows reference to this section

Three-dimensional modeling of indoor environments has a variety of
applications such as training and simulation for disaster management,
virtual heritage conservation, and mapping of hazardous sites. Manual
construction of these digital models can be time-consuming, and as a
result, automated 3D site modeling has garnered much interest in
recent years.

The first step in automated 3D modeling is the physical scanning of an
environment's geometry. An indoor modeling system must recover its own
poses within an environment while simultaneously reconstructing the 3D
structure of the environment itself \cite{chen2010indoor,
  liu2010indoor, kua2012loopclosure}. This is known as the
simultaneous localization and mapping (SLAM) problem, and is generally
solved by taking readings from laser range scanners, cameras, and
inertial measurement units (IMUs) at multiple locations within the
environment.

In this paper, we work with data obtained from a backpack-mounted
system, carried by an ambulatory human. Such a system provides
advantages over more common wheel-mounted systems in terms of agility
and portability, but results in higher difficulty and error when
performing localization \cite{liu2010indoor}. As a result, approaches
to texture mapping that rely on accurate camera poses produce poor
results, leading to the development of the methods contained in this
paper. Besides localization information and images, environment
geometry is required as well for texture mapping. In this paper we
work with low-resolution models obtained by fitting planar surfaces to
point clouds generated by the backpack system \cite{sanchez2012point}.

SOME CONCLUSION TO THE INTRODUCTION GOES HERE Though our texture
mapping procedure was designed with the aforementioned system in mind,
it is applicable to other systems, provided the required inputs of
planar geometry, images with rough extrinsic pose estimates, and
either known or estimated intrinsic camera calibration matrices. An
important benefit of our method is its relatively low complexity and
runtime, and independence of many steps, allowing for efficient tuning
of parameters and easy accomodation of new input data given updates in
localization or model-generation algorithms.

The remainder of the paper is organized as follows. Section
\ref{sec:simpleTextureMapping} explains how input images are projected
onto our geometry and a simple approach towards texturing. Section
\ref{sec:existingApproaches} covers more sophisticated existing
approaches, using image matching and stitching, and their performance
on our datasets. Section \ref{sec:2dAlignment} contains our approach
towards image alignment, followed by Section
\ref{sec:imageCompositing}, which describes our two methods of
compositing images. Section \ref{sec:resultsAndConclusions} contains
results and conclusions.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Background}
\label{sec:background}
\subsection{Existing Approaches}
\label{sec:existingApproaches}
\subsection{

@@

  The geometry of the texture mapping process for a planar surface is
  shown in Figure \ref{fig:projection}.  As described earlier, we are
  provided with a set of $M$ images to texture a target plane. Each
  image has a camera matrix $P_i$ for $i=1..M$, which translates a 3D
  point in the world coordinate system to a 2D point or pixel in image
  $i$'s coordinates. A camera matrix $P_i$ is composed of the camera's
  intrinsic parameters, containing focal length and image center, as
  well as extrinsic parameters which specify the rotation and
  translation of the camera's position in 3D world coordinates at the
  time that image $i$ is taken. These extrinsic parameters are
  determined by the backpack hardware and localization algorithms
  \cite{chen2010indoor, liu2010indoor, kua2012loopclosure} and are
  quite noisy.

@@

  \begin{figure}
    \begin{minipage}[b]{0.45\linewidth}
      \centering
      \includegraphics[height=1.5in]{Projection.pdf}
      \caption{Planes are specified in 3D space by four corners $C_1$
        to $C_4$. Images are related to each plane through the camera
        matrics $P_{1..m}$. }
      \label{fig:projection}
    \end{minipage}
    \hspace{0.5cm}
    \begin{minipage}[b]{0.45\linewidth}
      \centering
      \includegraphics[height=1.125in]{scoringFunction.jpg}
      \caption{We minimize camera angle $\alpha$ and distance $d$ by
        maximizing the scoring function $\frac{1}{d} (-1 \cdot
        \vec{c}) \cdot \vec{n}$}
      \label{fig:scoringFunction}
    \end{minipage}
  \end{figure}

  \subsection{Tile-based Texture Mapping}
  \label{sec:tileBasedMapping}
  Ignoring the fact that the camera matrices $P_{1..M}$ are
  inaccurate, a simple texturing approach can be performed by
  discretizing the target plane into small square tiles, in our case 5
  pixels across, and choosing an image to texture each tile.

  We choose to work with rectangular units to ensure that borders
  between any two distinct images in the final texture are either
  horizontal or vertical. Since most strong environmental features
  inside buildings are horizontal or vertical, any visible seams in
  our texture thus intersect them minimally and are less noticeable.

  In order to select an image for texturing a tile $t$, we must first
  gather a list of candidate images that contain all four of its
  corners, which we can quickly check by projecting $t$ into each
  image using the $P_i$ camera matrices. Furthermore, each candidate
  image must have been taken at a time when its camera had a clear
  line-of-sight to $t$, which can be calculated using standard
  ray-polygon intersection tests between the camera location, $t$, and
  every other plane \cite{rayintersection}.

  Once we have a list of candidate images for $t$, we define a scoring
  function in order to objectively select the best image for texturing
  $t$. Since resolution decreases and camera pose errors compound over
  distance, we wish to minimize the distance between cameras and the
  surfaces they texture. Additionally, we desire images that are
  projected perpendicularly onto the plane, maximizing the resolution
  and amount of useful texture available in their projections, as well
  as minimizing any parallax effects due to real-world geometry not
  accurately represented by our digital model. In other words, we wish
  to minimize the angle between the tile's normal vector and the
  camera axis for images selected for texture mapping. These two
  criteria can be met by maximizing the function $\frac{1}{d} (-1
  \cdot \vec{c}) \cdot \vec{n}$ as shown in Figure
  \ref{fig:scoringFunction}. Specifically, $d$ is the distance between
  the centers of a camera and a tile, and $\vec{n}$ and $\vec{c}$ are
  the directions of the plane's normal and the camera axis
  respectively.

@@

As Figure \ref{fig:compareAll}(a) demonstrates, this approach leads to
the best texture for each tile independently, but results in many
image boundaries with abrupt discontinuities between tiles, due to
significant misalignment between images. Given our high number of
input images, a better image selection procedure could be used, as
described in Section \ref{sec:imageCompositing}, but a more
significant and reliable improvement can be found through the
alignment of our images.

\section{Existing Approaches to Image Alignment}
\label{sec:existingApproaches}
Stitching together multiple images to produce a larger image is a
commonly performed task, with many successful approaches over the past
few decades. Parts of images are first matched to eachother, through
direct pixel comparisons, or more commonly through feature detection
and matching. Images are then adjusted to maximize matches, either by
calculating homographies between pairs of images, or by modifying
camera poses in 1 to 6 degrees of freedom.

Feature detection and matching works best when multiple unique visual
references exist in the environment and are present within multiple
images. In contrast, our indoor environments have a high prevalence of
bare surfaces, as well as repeating textures that cause difficulty in
disambiguating features. This lack of reliable reference points often
results in errors when matching images together.

Additionally, our datasets often contain long chains of images, which
leads to error accumulation when image correspondences are not
accurate. For example, when matching a long chain of images through
homographies, a pixel in the $nth$ image is translated into the first
image's coordinates by multiplying by the $3\times3$ matrices $H_1 H_2
H_3 ... H_n$. Errors in any of these homography matrices are
propagated to all further images, resulting in drift. Figure
\ref{fig:mosaic3D}(a) shows the output of the AutoStitch software
package, which performs homography-based image mosaicing
\cite{autostitch}. Though AutoStitch performs well in areas with dense
features, it has difficulty with even short segments of bare walls,
and produces errors that cause straight lines to appear
distorted. Many areas with fewer visual features or repeating texture
patterns failed outright using AutoStitch.

@@

Image projections can also be aligned by iteratively adjusting camera
poses to maximize matches. This is generally done over 6 degrees of
freedom \cite{liu2010indoor}. The approach in [2] works well locally,
but over larger areas results again in error accumulation and drift,
as shown in Figure \ref{fig:mosaic3D}(b). Furthermore, it adjusts
image locations using a long iterative process, which results in high
runtime.

\section{2D Image Alignment with Geometry Information}
\label{sec:2dAlignment}
In this section, we describe our method of efficient and robust image
alignment. Our approach consists of three main parts. First, all
images are projected onto the surface and lines within these
projections are detected. These lines are then matched up with
geometry-based lines comprising the surface's boundary and
intersection with other surfaces. Second, occlusion checks are
performed to remove invalid parts of each image for the target
surface. These two steps are image-independent and performed in
parallel. Third, we detect SIFT feature matches between pairs of
images and solve a linear least squares problem in 2D to maximize
matches.

All of our following calculations and alignments are performed in 2D,
partly for efficiency, partly because we have a large number of images
to choose from, and partly because the nature of our input data is
such that localization error chiefly occurs in two dimensions, in the
same plane as the surface being projected onto. Recall that our data
acquisition system is backpack-mounted, with cameras facing to the
sides of the operator. The operator makes efforts to walk parallel to
walls, and the localization and model-generation algorithms that
provide our input conform the operator's path to be straight and
parallel to detected surfaces \cite{kua2012loopclosure,
  sanchez2012point}. Since the operator walks in as upright a position
as possible, errors in roll, yaw, and the operator's distance from
parallel surfaces are minimal. Thus, our highest errors stem from
uncertainty in the operator's pitch, equivalent to rotation around the
camera axes, as well as location along surfaces. These equate to 2D
rotation and translation in a surface's plane.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\begin{abstract}
  Automated 3D modeling of building interiors is used in applications
  such as virtual reality and environment mapping. Texturing these
  models allows for photo-realistic visualizations of the data
  collected by such modeling systems. We perform these tasks using a
  backpack-mounted data acquisition system. Camera poses obtained by
  this system and used for texturing often suffer from inaccuracies
  however, resulting in visible discontinuities when successive images
  are projected onto a surface for texturing. Existing methods to
  stitch images together are often computationally expensive and work
  independently of pose estimates and geometry information. By
  intelligently selecting images for texturing, we develop
  an efficient method to adjust camera poses in 2D, followed by two
  different methods to composite images together, based on the
  geometry of surfaces being textured. The effectiveness of our
  method is demonstrated on a number of different indoor
  environments.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:introduction} % \label{} allows reference to this section
Three-dimensional modeling of indoor environments has a variety of
applications in fields such as architecture, conservation, and
simulation. Applying accurate textures to these models is important to
add context to digital visualizations, and allows for image-based
analysis of environments.

In this paper, we perform data acquisition with a backpack-mounted
system carried by an ambulatory human \cite{liu2010indoor}. This
backpack system contains a series of cameras, laser scanners, and
inertial measurement units, used to solve the simultaneous
localization and mapping (SLAM) problem, which produces backpack (and
therefore camera) poses over time as well as a point cloud
representing the surrounding environment \cite{chen2010indoor,
  liu2010indoor, kua2012loopclosure}. For environment geometry, we
work with models obtained by fitting low-resolution planar surfaces to
these point clouds \cite{sanchez2012point}.

While a human-carried system provides advantages over more common
wheeled systems in terms of agility and portability, it often leads to
relatively high localization error even after applying sophisticated
localization techniques \cite{liu2010indoor}. Furthermore, our model
geometry usually contains some error as well, and is often tuned to
ignore low-resolution features, such as small walls, furniture, and
other minor non-planar features commonly found in indoor
environments. As a result, methods for texture mapping that rely on
accurate camera poses and projection surfaces generally produce poor
results, while more sophisticated methods to accurately align images
in 3D lead to high runtime on our typically very large datasets. In
contrast, we present a method, which given a high number of images,
searches for a subset of them that can be successfully aligned with 2D
adjustments alone.

The remainder of the paper is organized as follows. Section
\ref{sec:simpleTextureMapping} explains how images are projected onto
our geometry and explains a simple approach towards texturing. Section
\ref{sec:existingApproaches} covers more sophisticated existing
approaches, and their performance on our datasets. Section
\ref{sec:2dAlignment} contains our efficient approach towards image
alignment, followed by Section \ref{sec:imageCompositing}, which
describes our two methods of compositing images. Section
\ref{sec:resultsAndConclusions} contains results and conclusions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The geometry of the texture mapping process for a planar surface is
shown in Figure \ref{fig:projection}.  We are provided with a set of
$M$ images to texture a target plane. Each image has a camera matrix
$P_i$ for $i=1..M$, which translates a 3D point in the world
coordinate system to a 2D point or pixel in image $i$'s coordinates. A
camera matrix $P_i$ is composed of the camera's intrinsic parameters,
containing focal length and image center, as well as extrinsic
parameters which specify the rotation and translation of the camera's
position in 3D world coordinates at the time that image $i$ is
taken. These extrinsic parameters are determined by the backpack
hardware and localization algorithms \cite{chen2010indoor,
  liu2010indoor, kua2012loopclosure} and are substantially noisy.

@@

\begin{figure}
  \begin{minipage}[b]{0.45\linewidth}
    \centering
    \includegraphics[height=1.5in]{Projection.pdf}
    \caption{Surfaces to be textured are specified in 3D space by
      corners $C_i$. Images are related to each plane through the
      camera matrics $P_{1..m}$. }
    \label{fig:projection}
  \end{minipage}
  \hspace{0.5cm}
  \begin{minipage}[b]{0.45\linewidth}
    \centering
    \includegraphics[height=1.125in]{scoringFunction.jpg}
    \caption{We minimize camera angle $\alpha$ and distance $d$ by
      maximizing the scoring function $\frac{1}{d} (-1 \cdot \vec{c})
      \cdot \vec{n}$}
    \label{fig:scoringFunction}
  \end{minipage}
\end{figure}

\subsection{Tile-based Texture Mapping}
\label{sec:tileBasedMapping}
Ignoring the fact that the camera matrices $P_{1..M}$ are inaccurate,
a simple texturing approach can be performed by discretizing the
target plane into small square tiles, in our case 5 pixels across, and
choosing an image to texture each tile directly.

We choose to work with rectangular units to ensure that borders
between any two distinct images in the final texture are either
horizontal or vertical. Since most strong environmental features
inside buildings are horizontal or vertical, any visible seams in our
texture will intersect them minimally and be less noticeable.

In order to select an image for texturing a tile $t$, we first gather
a list of candidate images that contain all four of its corners, which
we can quickly check by projecting $t$ into each image using the $P_i$
camera matrices. Furthermore, each candidate image must have been
taken at a time when its camera had a clear line-of-sight to $t$,
which can be calculated using standard ray-polygon intersection tests
between the camera location, $t$, and every other plane
\cite{rayintersection}.

Once we have a list of candidate images for $t$, we define a scoring
function in order to objectively select the best image for texturing
$t$. Since resolution decreases and camera pose errors become more
extreme over distance, we wish to minimize the distance between
cameras and the surfaces they texture. Additionally, we desire images
that are projected perpendicularly onto the plane, maximizing the
resolution and amount of useful texture available in their
projections, as well as minimizing any parallax effects due to
real-world geometry not accurately represented by our digital
model. In other words, we wish to minimize the angle between the
tile's normal vector and the camera axis for images selected for
texturing that tile. These two criteria can be met by maximizing the
function $\frac{1}{d} (-1 \cdot \vec{c}) \cdot \vec{n}$ as shown in
Figure \ref{fig:scoringFunction}. Specifically, $d$ is the distance
between the centers of a camera and a tile, and $\vec{n}$ and
$\vec{c}$ are the directions of the plane's normal and the camera axis
respectively.

@@

As Figure \ref{fig:compareAll}(a) demonstrates, this approach seleccts
images with high resolution, and handles parallax effects well, as
seen with the table and the poster at the lower-right of the
image. Obviously, there are many image boundaries with abrupt
discontinuities between tiles, due to significant misalignment between
images, though most of it appears to be reconcilable using 2d
transforms. Given our high number of input images, a better image
selection procedure will further improve results, as described in
Section \ref{sec:imageCompositing}, but a more significant and
reliable improvement can be found through first improving the
alignment of our images.

\section{Existing Approaches to Image Alignment}
Stitching together multiple images to produce a larger, seamless image
is a commonly performed task, with many successful approaches over the
past years. Generally, parts of images are matched to each other,
usually through direct comparisons or feature detection and
matching. Images are then transformed to maximize matches, often by
calculating homographies between pairs of images, or by iteratively
adjusting camera poses in 1 to 6 degrees of freedom.

Feature matching has a number of advantages over direct matching that
make it more suitable for our often non-planar data, and rotational
differences, as covered in X. Feature matching however, works best
when multiple unique visual references exist in the environment that
can be detected in multiple images. In contrast, our indoor
environments have a high prevalence of bare surfaces, as well as
repeating textures, such as similar windows, doors, and wall-mounted
decorations, that cause difficulty in disambiguating features. This
lack of reliable reference points often results in errors when
matching images together.

Additionally, our datasets often contain long chains of images, which
leads to error accumulation when image correspondences are not exact
and are compounded together. For example, when matching a long chain
of images through homographies, a pixel in the $nth$ image is
translated into the first image's coordinates by multiplying by the
$3\times3$ matrices $H_1 H_2 H_3 ... H_n$. Errors in any of these
homography matrices are propagated to all further images, resulting in
drift.

@@

Past work with our data acquisition backpack integrated image
stitching with an iterative localization algorithm, and performed
homography-based camera pose refinement. When run on long chains of
images, especially where features are sparse, it produces distorted
textures, as seen in Figure X, due to the problems described
above. This approach is also WHATEVER NONLINEAR, and its iterative
camera adjustment process over our large datasets is not guaranteed to
SOMETHING, and leads to prohibitively long computation time.

The AutoStitch software package performs homography-based alignment as
well, with additional provisions that attempt to reduce drift (cite
X,X). Though AutoStitch also performs well in areas with dense
features, it can not handle areas without features, and has trouble
aligning wall sections with even short segments of bare texture. The
example in Figure X was generated after manual tuning, and areas with
fewer visual features or repeating texture patterns simply failed
outright with AutoStitch.

\section{2D Image Alignment with Geometry Information}
\label{sec:2dAlignment}

In this section, we describe our method of efficient and robust image
alignment. Though state-of-the-art techniques towards image stitching
work in 3D, we sacrifice THAT WORD for efficiency, and elect to
perform image alignments entirely in 2D. Besides the reduced
complexity, we have found 2D alignment to work well for two main
reasons. First, because we have such a large number of images to
choose from, we have found that the compositing approache in Section X
is able to select a subset of images that align well, given only 2D
alignments. Furthermore, the nature of our input data is such that
localization error chiefly occurs in two dimensions, in the same plane
as the surface being projected onto. Recall that our data acquisition
system is backpack-mounted, with cameras facing to the sides of the
operator. The operator makes efforts to walk parallel to walls, and
the localization and model-generation algorithms that provide our
input conform the operator's path to be straight and parallel to
detected surfaces \cite{kua2012loopclosure, sanchez2012point}. Since
the operator walks in as upright a position as possible, errors in
roll, yaw, and the operator's distance from parallel surfaces are
minimal. Thus, our highest errors stem from uncertainty in the
operator's pitch, equivalent to rotation around the camera axes, as
well as location along surfaces. These equate to 2D rotation and
translation in a surface's plane.

Our alignment procedure consists of three main parts. First, all
images are projected onto the surface and lines within these
projections are detected. These lines are then matched up with
geometry-based lines comprising the surface's boundary and
intersection with other surfaces. Second, occlusion checks are
performed to remove invalid parts of each image for the target
surface. These two steps are image-independent and performed in
parallel. Third, we detect SIFT feature matches between pairs of
images and solve a linear least squares problem in 2D to maximize
matches.
